---
title: 'Introduction to latent Dirichlet allocation (LDA) in text analysis'
author: "A. Paxton (*University of Connecticut*)"
output:
  html_document:
    keep_md: yes
    number_sections: yes
---

Latent Dirichlet allocation (LDA) is a form of *topic modeling*, a class
of unsupervised machine learning methods that can be used to identify a
researcher-defined number of clusters of words within text data. LDA assumes that
each *document* (or meaningful "chunk" of text) made up of a mix of one or more *topics*;
each topic is made up of a mix of different words. The process looks at the
probabilistic co-occurrence of these terms in order to identify the topics, and 
researchers can provide more human-interpretable labels for each topic by looking at
the words that comprise them. Researchers can then explore how these topics 
occur across different documents.

To perform LDA, you'll need to have a series of texts from a number of different
documents. (As with other text analysis, what you consider a "document" will be
flexible: It could be abstracts, chapters, books, blog posts... whatever you
count as a meaningful "chunk" of text to be analyzed.)

In this exercise, we'll walk through the process of performing LDA in R.

***

# Preliminaries

First, we'll need to prepare for our exercises. We'll do this by loading the
packages we need, including---if needed---installing those packages. For the
sake of the exercise, we'll load the packages silently, meaning that we won't
clutter our R markdown output with a bunch of warnings and output messages
from the loading and installation process.

```{r clear-workspace}

# clear the workspace (useful if we're not knitting)
rm(list=ls())

```

```{r function-check-for-packages, include=FALSE}

# make sure we can load packages 
# (thanks to https://gist.github.com/smithdanielle/9913897)
load_or_install_packages <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg)) {
    install.packages(new.pkg, 
                     dependencies = TRUE,
                     repos="http://cloud.r-project.org/")}
  sapply(pkg, require, character.only = TRUE)
}

```

```{r load-packages, message=FALSE, results="hide"}

# specify which packages we'll need
required_packages = c("tidyverse",
                      "topicmodels",
                      "tidytext")

# install them (if necessary) and load them
load_or_install_packages(required_packages)

```

***

